@$\lambda$: Conservatively Extending a Type System From Within

### 2014-02-27 att-icfp14/att-icfp14.tex
Exposition of sec 1 and 2

@@

\maketitle
\begin{abstract}
Researchers often extend an existing language with new type and operator constructors. 
But because this is not generally possible from within the language, researchers must implement extensions as distinct forks of the language.
This limits their utility: building applications where different components are written in languages with different type systems is unsafe and unnatural. 
An {internally-extensible} type system could help address these issues, but designing an extension mechanism that is both expressive and reliable is a challenge: extensions must not be permitted to compromise key metatheoretic properties of the language, nor should they be allowed to interfere with one another, in any combination. This latter property, \emph{conservativity}, is not guaranteed by prior mechanisms.

We introduce @$\lambda$, a simply-typed lambda calculus with simply-kinded type-level computation where the  semantics of new type and operator constructors can be defined in a functional style from within. Its semantics are structured as an extensible variant of the Harper-Stone elaboration semantics used to specify Standard ML. We incorporate typed compilation techniques and a form of type abstraction that together allow us to prove type safety, decidability and conservativity. We intend @$\lambda$ to serve as a minimal foundation for future work on extension mechanisms for typed programming languages, but it is already quite expressive. We begin by showing how a more natural concrete syntax can be introduced by a type-directed desugaring to @$\lambda$. Then, we show how a number of abstractions that must be built into other languages can be implemented orthogonally, as libraries, within @$\lambda$. We conclude with a discussion of its present limitations.  
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}
Typed programming languages are typically organized around a collection of indexed type and operator constructors. The simply-typed lambda calculus (STLC), for example, provides a single type constructor, $\rightarrow$, indexed by a pair of types, and two operator constructors, $\lambda$, indexed by a type, and $\opvar{ap}$, which can be thought of as being indexed trivially. We can extend it with universally quantified and recursive types by adding the type constructors $\forall$ and $\mu$, each indexed by a type binder (a type-level term that may refer to a free type variable), and their associated operator constructors. It is tempting to stop here: the language is now universal \todo{right terminology?} (i.e. Turing complete) and the dynamic semantics of operators associated with other useful types, including sums and products, can be defined by a method analagous to Church encodings in the untyped lambda calculus \cite{reynolds}. This is of clear theoretical and pedagogical interest. But this calculus is plainly impractical as a programming language. Reynolds, striking a cautionary tone reminiscent of Perlis in his description of the ``Turing tarpit'' \cite{perlis}, put it succinctly \cite{reynolds}: 
\begin{quote}
To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style.
\end{quote}

A simple  language suffices when only the dynamic semantics of a program are being considered but continuing to add type and operator constructors (collectively, \emph{constructs}) to a language can make reasoning about programs simpler and more precise and support a more natural programming style. Consistent with this view, typed programming languages generally expose more constructs externally than their compilers work with internally.  
A functional language like ML, for example, might expose $n$-ary product types (tuples), labeled product types (records), recursive labeled sum types (datatypes) and forms of type abstraction (polymorphism and abstract data types). A compiler for the language might then translate these to an intermediate  language (IL) with only simple products and sums for further compilation to machine code. Correctly implementing the dynamic semantics of the language is the primary concern during this phase, so a simple IL that decreases the number of cases that must be considered when implementing and verifying the compiler is a wise choice.

It is again tempting to stop here. Indeed, languages like ML are often referred to as ``general-purpose languages'' and have been used successfully for a range of computing tasks. The constructs they provide certainly occupy ``sweet spots'' in the design space, and many abstractions can be encoded using these constructs in a completely satisfactory manner. However, they admit no new ``universality'' theorems  stronger than those for the simple calculus described above, and situations continue to arise in both research and practice where new constructs are desirable:

\begin{enumerate}
\item General-purpose constructs continue to evolve. There are many  variations on record types, for example: structurally-typed records, or records with forms of dynamic dispatch. We will show an example of records with  prototypic dispatch in Sec. \ref{prototypes}, and argue that they cannot be safely and naturally defined in terms of standard functional constructs. Sum types also admit variants, as we will discuss in the next section.

\item Specialized type systems that enforce stronger invariants than general-purpose abstraction mechanisms can enforce are often proposed by researchers. One need not take more than a cursory glance through the literature to discover type systems for parallel programming \cite{a,b}, concurrent programming \cite{cml,a,b}, distributed programming \cite{tom7}, dataflow programming \cite{reactiveml}, authenticated data structures \cite{popl13}, information flow \cite{walker00, smith2001}, databases \cite{db}, aliasing \cite{naden12}, effects \cite{effects},  network protocols \cite{sekar99}, regular expressions \cite{regex}, units of measure \cite{keneddy} and many others.
\item Interoperability layers that allow programmers to safely and naturally interact with libraries written in a different language require enforcing the type system of that language within the calling language. Using library-based interoperability layers today can lead to safety issues when this is not possible.
%\item General-purpose abstraction mechanisms are typically implemented uniformly. For example, inductive datatypes are typically implemented using tagged pointers to a payload. In some cases (e.g. natural numbers) it may be desirable to implement an abstraction in an alternative manner (e.g. as an integer) so that certain operations perform better. Parallel abstractions are of particular note, as they can benefit substantially from specialized implementation strategies on specialized hardware (e.g. a GPU).  Note that if control over internal implementation details were the only relevant use case, abstract datatypes might be appropriate.
\end{enumerate}

Today, complete solutions to these problems generally take the form of a \emph{fork} of an existing general-purpose language (discussed further in the next section). This is unsatisfying because then these solutions are  \emph{non-orthogonal}: a programmer can choose either a language supporting a reasonable approach to parallel programming  or one that builds in support for working with and statically reasoning about regular expressions, but there is not necessarily a language that supports both. Even if a separation of such concerns can be achieved, such that each component of an application can be written using a specialized language (as advocated by proponents of the \emph{language-oriented approach} to software development \cite{language-oriented}), there will be an \emph{interoperability} problem at the interface between components, as described in item 3 above.

>>>>

\lefthyphenmin=6

@@

\maketitle
\begin{abstract}
Researchers often need to extend an existing language with new type and operator constructors to realize a new abstraction in its strongest form. 
But this is not generally possible from within, so it is common to develop dialects and ``toy'' languages.
Unfortunately, taking this approach limits the utility of these abstractions: they cannot be imported by clients of other dialects and languages, and building applications where different components rely on different type systems is both unsafe and unnatural. 
%An {internally} {extensible} type system could help address these issues, but designing an extension mechanism that is both expressive and safe is non-trivial: extensions must not be permitted to compromise key metatheoretic properties of the language, nor should they be allowed to weaken one another, in any combination.

We introduce @$\lambda$, a simply-typed lambda calculus with simply-kinded type-level computation where new indexed type and operator constructors can be declared  from within. %It is structured in four layers: at the top-level, extension providers declare new indexed type constructors and term-level operator constructors. 
Type-level functions associated with operator constructors define their static and dynamic semantics, the latter by translation to a fixed typed internal language. By lifting compiler correctness techniques pioneered by the TIL compiler for Standard ML into the language, the ``actively typed semantics'' guarantees type safety. Going further, the semantics enforce a novel form of representation independence at extension boundaries that ensures that extensions are mutually conservative (i.e. they do not weaken or interfere with one another, so that they can  be used together in any combination). 
%Totality of the type-level language guarantees decidability of type checking and equality constraints on type indices makes type equivalence decidable. 
We intend @$\lambda$ as a minimal foundation for future work on safe language-integrated extension mechanisms for typed programming languages, but it is already quite expressive. We demonstrate by showing how a conventional concrete syntax can be introduced by type-directed dispatch to Core @$\lambda$, then discuss a number of typed language fragments that can be introduced as orthogonal ``libraries'' (and discuss the sorts of fragments that, as yet, cannot).
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}
Typed programming languages are often described in fragments, each consisting of a small number of type constructors (often one) and associated (term-level) operator constructors. The simply typed lambda calculus (STLC), for example, consists of a single fragment containing a single type constructor, $\rightarrow$, indexed by a pair of types, and two operator constructors: $\lambda$, indexed by a type, and \verb|ap|, which we may think of as being indexed trivially. A fragment is often identified by the type constructor it is organized around, so the STLC might also be called $\mathcal{L}\{\rightarrow\}$, following the notational convention used by Harper \cite{pfpl} and others.

G\"odel's $\mathbf{T}$ consists of the $\rightarrow$ fragment and the \verb|nat| fragment, defining natural numbers and a recursor that allows one to ``fold'' over a natural number. We might thus call it $\mathcal{L}\{\rightarrow \mathtt{nat}\}$. This language is  more powerful than the STLC because the STLC admits an embedding (trivially) into  $\mathbf{T}$ but the reverse is not true. Buoyed by this fact, we might go on by adding fragments that define sums, products and various forms of inductive types (e.g. lists), or perhaps a general mechanism for defining inductive types. Each fragment clearly increases the expressiveness of our language.

If we consider the $\forall$ fragment, however, defining universal quantification over types (i.e. parametric polymorphism), we might take pause. In $\mathcal{L}\{\rightarrow\,\forall\}$, studied variously by Girard as System $\mathbf{F}$ \cite{girardF} and Reynolds as the polymorphic lambda calculus \cite{polylam}, it is known that sums, products, and inductive and co-inductive types can all be weakly defined via a technique similar to Church encodings \cite{reynolds}\todo{what to cite for this?}. While of substantial theoretical (and pedagogical) interest, this fact does not render irrelevant the construction of a language like $\mathcal{L}\{\rightarrow\forall~\mathtt{nat} + \times\}$. Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, cautioned against using the existence of a weak encoding as an argument for the practicality of programming with the corresponding embedding directly \cite{Reynolds94anintroduction}: 
\begin{quote}
To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
\end{quote}

Finding a weak encoding of a language fragment in terms of a collection of others is useful if \emph{implementing the dynamic semantics correctly} is one's main interest, but adding new fragments directly to a language can make statically reasoning about programs more precise, support a more natural programming style and endow the language with a more favorable cost semantics.  
Consistent with this view, typed programming languages like ML expose, for example, both $n$-ary product types and record types, despite the fact that any particular product or record type is weakly definable using  polymorphic functions (or simply in terms of the other). The compiler for such a language, on the other hand, \emph{is} mainly concerned with implementing the dynamic semantics correctly, and indeed, compilers often use the fact that weak definability results are quite readily derivable to their advantage, translating the constructs of their external language (EL) to those of a much simpler internal language (IL) after typechecking is complete. %Correctly implementing the dynamic semantics of the language becomes the primary concern during this phase, so a simple IL that decreases the number of cases needing consideration when implementing and verifying the compiler is a wise choice.

Having established why a minimal language like $\mathcal{L}\{\rightarrow\forall\}$, while suitable as an internal language, needs to be extended with additional fragments before it is suitable for use as a human-facing external language, we might now ask whether modern ``general-purpose'' languages like ML are fundamentally different. These languages have been used successfully for a wide variety of computing tasks, admitting many more strong encodings than a  calculus like $\mathcal{L}\{\rightarrow \forall\}$. It is certainly not our intention to argue against the notion that the particular fragments they have chosen to privilege occupy ``sweet spots'' in the design space. But in fact, situations continue to arise in both research and practice where the direct introduction of a new fragment is desirable because it can still only be weakly defined in terms of existing fragments, or a strong encoding is possible but impractically verbose or inefficient:
\begin{enumerate}
\item General-purpose abstractions continue to evolve and admit variants. There are many  variations on record types, for example: records with functional record update, prototypic delegation or specified field ordering (that is, labeled tuples) are either awkward to strongly define or cannot be. Similarly, sum types also admit variants (e.g. various forms of open sums). Even something as seemingly simple as a type-safe \verb|printf| operator requires extending a language like Ocaml  \cite{ocaml-printf}.

\item Perhaps more interestingly, specialized type systems that enforce stronger invariants than general-purpose constructs are capable of enforcing are often developed by researchers. One need take only a brief  excursion through the literature to discover language extensions that support parallel programming \cite{a}\todo{citations}, concurrent programming \cite{cml}, distributed programming \cite{tom7}, dataflow programming \cite{reactiveml}, authenticated data structures \cite{popl13}, information flow security \cite{walker00, smith2001}, database queries \cite{db}, aliased references \cite{naden12}, network protocols \cite{sekar99}, units of measure \cite{keneddy} and a bounty of others.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
\item Foreign function interfaces (FFIs) that allow programmers to safely and naturally interact with libraries written in an existing language require enforcing the type system of the foreign language within the calling language. %Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe. 
Safe FFIs generally require direct extensions to the language. For example, MLj extended Standard ML with constructs for safely and naturally interfacing with Java \cite{mlj}. For any other language, including  others on the JVM like Scala and the many languages that might be accessible via Java's native FFI, there is no way to guarantee that language-specific invariants are statically maintained, and the interface is certainly far from natural.
\end{enumerate}

These language fragments are generally disseminated as distinct \emph{dialects} of an existing general-purpose language, constructed either as a fork, a feature behind a compiler flag or, using tools like compiler generators, DSL frame\-works or  language workbenches. This, we argue, is quite unsatisfying: a programmer can choose either a dialect supporting an innovative approach to parallel programming or one that builds in support for statically reasoning about units of measure, but there may not be an available dialect supporting both. Forming such a dialect is alarmingly non-trivial, even in the rare situation where a common framework has been used or the dialects are implemented within the same compiler, as these mechanisms do not guarantee that different combinations of individually sound language fragments remain sound when combined. Metatheoretic and compiler correctness results can only be derived for the dialect \emph{resulting} from language   composition operations, so even  fragmentary frameworks induce an unbounded, combinatoric collection of proof obligations that providers have little choice but to  burden clients with.% We will consider this further in our discussion of related work in Sec. \ref{related-work}.

These are not the sorts of problems usually faced by library providers. Well-designed languages preclude the  possibility of conflict between two libraries and ensure that the semantics of one library  cannot be weakened by another by strictly enforcing abstraction barriers. For example, a module declaring an abstract type in ML can rely on its representation invariants no matter which other modules are in use, so clients can assume that they will operate robustly in combination without attending to burdensome ``link-time'' proof obligations.

We draw inspiration from this approach to design a mechanism that makes it possible to extend the static and dynamic semantics of an external portion of the language with new type and operator constructors from within. The semantics of the language as a whole constrains these extensions to maintain type safety and decidability of type checking and type equality. Moreover, an important class of lemmas that play a key role in  proofs that an extension enables strong encodings of a language fragment can be derived assuming a ``closed world'' (that is, in the completely traditional manner where one constructs a complete language and reasons inductively). The language  guarantees that they will be conserved in the  ``open world'' by enforcing  representation independence at extension boundaries.

Our mechanism essentially tasks the type-level language, in a controlled manner, with implementing portions of the logic that would normally be implemented in the front-end of the compiler. When typechecking a user-defined operator invocation, the semantics invoke a type-level function associated with the operator's constructor to \emph{compute} both a type and a translation to a fixed typed internal language (together called a \emph{derivate}). Language fragments that can be weakly defined in terms of the typed internal language can be made strongly definable in the external language by introducing, in this way, the necessary  constructors and static semantics. We call this an \emph{actively typed semantics}.

We will begin in Sec. \ref{atlam} by introducing a minimal calculus with an actively typed semantics, @$\lambda$. The external language builds in only the $\rightarrow$ fragment. We then use the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ and a fragment defining $n$-ary products as pedagogical examples to explain how extensions are written and motivate the constraints that the semantics must impose to make the guarantees that we claim. We then examine the semantics in greater detail and state and sketch proofs of strong safety and conservativity theorems in Sec. \ref{theory}. \todo{organization of rest of sections}

\section{@$\lambda$}
\begin{figure*}[t]
\small
$$\begin{array}{rccl}	
\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort} 
%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
\pipe e\\
		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\taut{def}} \pipe 
												\topp{\theta}{\theta}\\
\\
\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
%														\efix{x}{\tau}{e} \pipe 
														\elam{\evar{x}}{\tau}{e} \pipe 
														\eop{Tycon}{op}{
															\tauidx
														}{
  												    		\splat{e}{1}{n}
														} \\
									& 		&		& 	\\

\textbf{internal terms} 				& 	\iota	&	::=	&	\evar{x} \pipe 
												\ifix{\evar{x}}{\sigma}{\iota} \pipe
												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
												\iapp{\iota_{1}}{\iota_{2}} \pipe
												\iintlit \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}} \\
										& & \pipe & 
												\iunit \pipe
												\ipair{\iota_{1}}{\iota_{2}} \pipe 
												\ifst{\iota} \pipe
												\isnd{\iota} \pipe
												 \iinl{\sigma_2}{\iota_1} \pipe \iinr{\sigma_1}{\iota_2} \pipe \icase{e}{x}{e_1}{x}{e_2} \\

%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe \dint \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} \pipe \dsum{\sigma_1}{\sigma_2}

\\
\\

\hspace{-5pt}\textbf{type-level terms} 	& \tau 	& ::= 	& 	\tvar{t} \pipe 
														\tlam{t}{\kappa}{\tau} \pipe 
														\tapp{\tau_1}{\tau_2} \pipe
														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
														\\

	 			& 		& \pipe	& 	 \iintlit \pipe \iop{\tau_1}{\tau_2} \pipe \tlabel{label} \pipe \tunit \pipe 
														\tpair{\tau_{1}}{\tau_{2}} \pipe 
														\tfst{\tau} \pipe 
														\tsnd{\tau} 
														\\	
									&       & \pipe & \tinl{\kappa_2}{\tau_1} \pipe \tinr{\kappa_1}{\tau_2} \pipe \tsumcase{\tau}{x}{\tau_1}{x}{\tau_2} \\
\text{equality}  & & \pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
														\\													\text{types} 						& 		& \pipe	& 	\ttypestd \pipe \tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}\\
						%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\

\text{derivates} 				& 		 & 	\pipe	&	\tden{\tauiterm}{\tautype}  \pipe \tderbind{x}{t}{\tau}{\tau_1}\\% \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
 %& & \pipe & 
%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
%														 \\

\text{reified IL}		&		&	\pipe	&	\titerm{\bar \iota} \pipe \titype{\bar \sigma} \\
% &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
% 												\trepof{\tau} \pipe \dup{\tau}\\
	& \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 
	%\pipe \ilam{x}{\bar \sigma}{\bar \iota} \pipe \iapp{\bar \iota_1}{\bar \iota_2} 
	\pipe \cdots \pipe \iup{\tau} \pipe \tvalof{\tauiterm}{\tautype} \\						
 & \bar{\sigma} & ::= & \darrow{\bar \sigma_1}{\bar \sigma_2} \pipe \cdots \pipe \dup{\tau} \pipe \trepof{\tau} \\
\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \kint \pipe
											    \klabel \pipe
												\kpair{\kappa_{1}}{\kappa_{2}} \pipe 
												\ksum{\kappa_1}{\kappa_2} \pipe
												\kITerm \pipe \kIType
\vspace{-10pt}
\caption{\small Syntax of Core \atlam. Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{Tycon}$ ranges over type constructor names, $\opvar{op}$ ranges over operator constructor names, $\iintlit$ ranges over integer literals, $\tlabel{label}$ ranges over label literals (see text) and $\oplus$ ranges over the standard total binary operations over integers (e.g. addition). The form $\tvalof{\tauiterm}{\tautype}$ is  used internally by the semantics (it need not be supported by a parser).%The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.
\vspace{-10pt}
\end{figure*}
\label{z}&\quad\tops{z}{\kunit}{i}{a}{\tlam{idx}{\kunit}{\tlam{args}{\klist{\kDen}}{
	\tapp{\tapp{\tvar{if\_empty}}{\tvar{args}}}{
	}}}};\\
\label{s}&\quad\tops{s}{\kunit}{i}{a}{\tlam{idx}{\kunit}{\tlam{args}{\klist{\kDen}}{\\
& \quad\quad
		\tapp{\tvar{pop\_final}}{\tvar{args}}}{\tlam{x}{\kITerm}{\tlam{t}{\kTypeBlur}{
	}}};\\
\label{rec}&\quad\tops{rec}{\kunit}{i}{args}{\tlam{idx}{\kunit}{\tlam{args}{\klist{\kDen}}{\\
& \quad\quad
	\tapp{\tapp{\tvar{pop}}{\tvar{args}}}{\tlam{x1}{\kITerm}{\tlam{t1}{\kTypeBlur}{\tlam{args}{\klist{\kDen}}{\\
	&\quad\quad\tapp{\tapp{\tvar{pop}}{\tvar{args}}}{\tlam{x2}{\kITerm}{\tlam{t2}{\kTypeBlur}{\tlam{args}{\klist{\kDen}}{\\
	&\quad\quad \tapp{\tapp{\tvar{pop\_final}}{\tvar{args}}}{\tlam{x3}{\kITerm}{\tlam{t3}{\kTypeBlur}{\\
	}}}}}}}}}}}}}}
&}{XXX}{i}{\tlam{idx}{\kunit}{\titype{\dint}}};\\
\label{nattype}&\etdef{nat}{\kTypeBlur}{\ttype{Nat}{\tunit}}{\\
& \elet{plus}{\ttype{Arrow}{(\tvar{nat}, \ttype{Arrow}{(\tvar{nat}, \tvar{nat})})}}{
	\elam{x}{\tvar{nat}}{
		\elam{y}{\tvar{nat}}{
			\\&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
			}}}
}{\\
& \elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{}}}}{\\
& \eop{Arrow}{ap}{\tunit}{plus; \eop{Arrow}{ap}{\tunit}{two; two}}
}}}
\vspace{-15px}
\caption{\small G\"odel's $\mathbf{T}$ in \atlam, used to calculate 2+2. The simple helper functions $\tvar{if\_empty}$, $\tvar{pop}$, $\tvar{pop\_final}$, $\tvar{check\_type}$ all have return kind $\ksum{\kDen}{\kunit}$ (see text). Their definitions can be found in the appendix. We use \textsf{let} to bind both external and type-level variables for clarity of presentation; these can be eliminated by manually performing the indicated substitutions (or added to the semantics).}
\label{nat}
%\begin{figure}[t]
%\begin{lstlisting}
%tycon Nat of 1 with  
%    schema $\lambda$idx:1.$\blacktriangledown$($\mathbb{Z}$) 
%    opcon Z of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].is_empty args $\llbracket$$\triangledown$(0) as Nat[()]$\rrbracket$)
%    opcon S of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].pop_final args $\lambda$x:ITm.$\lambda$ty:$\star$. 
%      checktype ty Nat[()] $\llbracket$$\triangledown$($\vartriangle$(x)+1) as Nat[()]$\rrbracket$)
%    opcon Rec of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].
%      pop args $\lambda$x1:ITm.$\lambda$ty1:$\star$.$\lambda$args':list[Elab].
%      pop args' $\lambda$x2:ITm.$\lambda$ty2:$\star$.$\lambda$args'':list[Elab].
%      pop_final args'' $\lambda$x3:ITm.$\lambda$ty3:$\star$.
%      check_type t1 Nat[()] (
%      check_type t3 Arrow[(Nat[()], Arrow[(t2, t2)])] 
%      	$\llbracket$$\triangledown$((fix f:$\mathbb{Z} \rightarrow$ rep(t2) is $\lambda$x:$\mathbb{Z}$.
%	        if x = 0 then $\vartriangle$(x2) else $\vartriangle$(x3) (x - 1) (f (x - 1))) $\vartriangle$(x1)) as t2$\rrbracket$))
%end
%\end{lstlisting}
%\caption{An implementation of primitive natural numbers as internal integers in @$\lambda$.}
%\label{nat-atlam}
%\end{figure}
In this section, we will develop an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation called @$\lambda$. More specifically, the level of types, $\tau$, will itself form a simply-typed lambda calculus. \emph{Kinds} classify type-level terms in the same way that types conventionally classify expressions. Types become just one kind  of type-level value (which we will write $\kTypeBlur$, though it is also variously written $\star$, \verb|T| and \verb|Type| in various settings). Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions. In the semantics for this calculus, our kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction allow us to prove strong type safety, decidability and conservativity theorems.

The syntax of @$\lambda$ is given in Fig. \ref{grammar}. An example of a program containing an extension that implements the static and dynamic semantics of G\"odel's \textbf{T} is given in Fig. \ref{nat}. We do not, of course, deny that natural numbers can be strongly encoded with a similar usage and asymptotic performance profile (but with potentially relevant additional function call overhead) by existing means (e.g. by an abstract type). We will provide more sophisticated examples where this is not the case below. %It consists of:
%a type constructor declaration, $\fvar{Nat}$, indexed trivially, together with three operator constructors, also all indexed trivially, that implement the standard introductory forms for natural numbers as well as the recursor operator (as in G\"odel's T \cite{pfpl}). Following the type constructor declaration, we apply $\fvar{Nat}$ with the trivial index, $\tunit$, to form the type $\tvar{nat}$. Finally, we write an external term that uses the operators associated with $\fvar{nat}$ and the built-in constructor $\fvar{Parr}$, governing partial functions, to define an addition function and compute the addition of the natural numbers  two and two. We will introduce a more convenient concrete syntax in later portions of this thesis; for now we will restrict ourselves to the abstract syntax so that this example can directly aid in understanding the semantics.

\subsection{Programs}

A program, $\rho$, consists of a series of static declarations followed by an external term, $e$. The syntax for external terms contains three core forms: variables, functions, and a form for all other operator invocations, which we will discuss below. Compiling a program consists of first \emph{kind checking} its static declarations and type-level terms then \emph{type checking} the external term and, simultaneously, \emph{translating} it to a term, $\iota$, in a \emph{typed internal language}. In our calculus, the internal language contains partial functions (via the generic fixpoint operator of Plotkin's PCF \cite{PCF}), simple product and sum types and a base type of integers (to make our example interesting and give a nod to practicality on contemporary machines). In practice, the internal language could be any intermediate language for which type safety and decidability of typechecking are known, as @$\lambda$ requires these to establish the corresponding theorems about the language as a whole.

The \emph{active typing judgement} relates an external term to a \emph{type} and an internal term, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$: 
$$\ecompilesX{e}{\tau}{\iota}$$

This judgement follows standard conventions for the typing judgement of a simply-typed lambda calculus in most respects. The typing context $\Gamma$ maps variables to types and obeys standard structural properties. The key novelties here are that the available type and operator constructors are not fixed by the language, but rather are tracked by the constructor context, $\fvalCtx$, and that a translation is simultaneously derived. The dynamic semantics of external terms are defined by their translation to the internal language. For this judgement, the two contexts and the external term can be thought of as input, while the type and translation are output (we do not consider a bidirectional approach in this work, though this may be a promising avenue for future research).

\subsection{Types}
User-defined type constructors are declared at the top-level of a program (or package, in a practical implementation) using \textsf{tycon}. Each type constructor in the program must have a unique name, written e.g. \fvar{Nat}. We do not intrinsically consider the issue of namespacing, under the assumption that globally unique names can be generated by some extrinsic mechanism (e.g. a URI-based scheme). A type constructor must also declare an \emph{index kind}, $\kappaidx$. Types themselves are type-level values of kind $\kTypeBlur$ and are introduced by applying a previously declared type constructor to a type-level term of its index kind, $\ttype{Tycon}{\tauidx}$. The kind $\kTypeBlur$ also has an elimination form: a type can be case analyzed against a type constructor in scope to extract its index. Like open datatypes, there is no longer a notion of exhaustiveness. To maintain totality, we require the default case.

To permit the implementation of interesting type systems, the type-level language includes several other kinds of data alongside types. We lift several  standard functional constructs to the type level: unit ($\kunit$), binary sums ($\ksum{\kappa_1}{\kappa_2}$), binary products ($\kpair{\kappa_1}{\kappa_2}$), lists ($\klist{\kappa}$) and integers ($\dint$). We also include labels ($\klabel$), written in a slanted font, e.g. $\tlabel{myLabel}$, which are atomic string-like values that can only be compared, here only for equality, and play a distinguished role in the expanded syntax, as we will later discuss. Our first example, $\fvar{Nat}$, is indexed trivially, i.e. by unit kind, $\kunit$, because there is only one natural number type, $\ttype{Nat}{\tunit}$, but we will show examples of type constructors that are indexed in more interesting ways in later portions of this work. For example, $\fvar{Tuple}$ is indexed by a list of types and $\fvar{LabeledTuple}$ is indexed by a list pairing labels with types.

Two type-level terms of kind $\kTypeBlur$ are equivalent if they apply the same constructor, identified by name, to equivalent indices. Going further, we ensure that deciding type equivalence requires only checking for syntactic equality after evaluation to normal form by imposing the restriction that type constructors can only be indexed by kinds for which equivalence can be decided in this way. All combinations of kinds introduced thus far  have this property and this is sufficient for many interesting examples. We leave  introducing a richer notion of  equivalence of types that  extension providers can extend as future work, as the metatheoretic guarantee that typing respects type equivalence would be quite a bit more complex in such a setting.% (a na\"ive approach to this would impose non-trivial extrinsic proof obligations onto extension developers that, unlike in others in this thesis, could threaten type safety).

Type constructors are not first-class; they do not themselves have arrow kind as in some kind systems (e.g.  \cite{Watkins08}\todo{cite}; Ch. 22 of \emph{PFPL} describes a related system \cite{pfpl}). The type-level language does, however, include total functions of conventional arrow kind, $\karrow{\kappa_1}{\kappa_2}$. Type constructor application can be wrapped in a type-level function to emulate first-class type constructors (and indeed, such a wrapper could be generated automatically, though we avoid this for simplicity in our semantics). Equivalence at arrow kind does not coincide with $\beta$-equivalence, so type-level functions cannot appear in type indices. Our treatment of equivalence in the type-level language is thus quite similar to the treatment of term-level equality using ``equality types'' in a language like Standard ML. Indeed, one might imagine that a practical implementation of our calculus would start with a pure, total subset of the term language of ML to construct the type-level language (consistent with its early development as a ``metalanguage'' and, as we will see, the role of our type-level language as a DSL for implementing type systems and translators, which functional languages are widely considered well-suited for as-is). We leave the development of such an implementation (and of bootstrapping type-level and internal languages that are themselves user-defined or extensible) as areas for future work.
\subsection{Operators}
User-defined operator constructors are declared using \textsf{opcon}. For reasons that we will discuss, our calculus associates every operator constructor with a type constructor. The \emph{fully-qualified name} of the operator constructor, e.g. $\fvar{Nat}.\opvar{z}$, must be unique. Operator constructors are indexed by type-level values, like type constructors, and so also declare an index kind $\kappaidx$. In our first example, all the operator constructors are indexed trivially, but later examples will use more interesting indices. For example, the projection operators for tuples and labeled tuples use numbers and labels as indices, respectively.
Note that neither operator constructors nor operators are first-class type-level values (we plan to investigate  lifting the latter restriction) and we do not impose any restrictions on their index kinds.
